# Intensive Care Medicine

## Approach to Critical Ill Patients

### Assesment of illness severity

Important for:

1.  resource allocation

2.  Hospital administrative policies

3.  asses quality of care

Two most commonly used scoring systems:

1.  SOFA (Sequential Organ Failure Assessment)

2.  APACHE (Acute Physiology and Chronic Health Evaluation)

#### SOFA Score

Includes 6 organ systems, with each graded 0 to 4 according to the degree of dysfunction. Increased score correlate with mortality and can be evaluated repeatedly [@harrison_2022].

A derivation, the qSOFA or quick SOFA intented to screen patients for srisk of poor outcomes from sepsis. It is not intended for sepsis diagnostic screening tool, but in it often used as such in resources poor area. qSOFA used for bedside evaluation that may identify patient with suspected infection who are at greater risk for a poor outcome outside the ICU. The score range from 0 to 3 points in each three category including blood preasure, respiratory rate and mental status. Poor outcimes predicted if there at least 2 clinical criteria: (1) respiratory rate $\geq$ 22/min, altered mental status, or systolic BP $\le$ 100 mmhg.

```{r}
#| echo: false
#| output: false

library(googlesheets4)
library(gt)

sofa <- read_sheet("https://docs.google.com/spreadsheets/d/1X1v0OOdzwSVROfs8C8JMvcjdi0EUFy9cQGwfcwFLuqE/edit#gid=0", sheet = 1)

sofa_tbl <- 
  gt(sofa) |> 
  tab_header(
    title = "Calculation of SOFA Score"
  ) |>
  tab_spanner(
    label = "Score",
    columns = c("0","1","2","3","4")
  )
```

```{r}
sofa_tbl
```

#### APACAHE Score

SLightly more complicated than SOFA Score. (updated later)

## Shock

### Initial Evaluation

shock is a multisystem end-organ hypoperfusion. The resulting hypoperfusion followed by tissue hypoxia with accompanying lactic acidosis. Clinical Indicator:

1.  Reduced MAP
2.  Tachycardia
3.  Tachypnea
4.  Cool Skin and Extremities
5.  Acute Altered mental Status
6.  Oliguria

Reduced MAP could be the product of decreased cardiac output and/or systemic vascular resistance (SVR). Thus every shock patients should be evaluated for adequacy of cardiac output. Sign of diminished cardiac output includes (Cold Shock):

-  A narrow pulse preassure (SBP - DBP), marker of stroke volume
-  Cool extremities and delayed capillary refill time (COLD SHOCK). Palpate proximal extremities (eg Thigh) rather than distal extremities to determine relative "coolness" as peripheral artery disease may always have cool distal extremities.

Contrary, there are sign of increased cardiac output (Warm Shock), that may results from decreased SVR:

-  A widened pulse pressure (particularly reduced DBP)
-  Warm extremities with bounding pulse,
-  Rapid capillary refill time

If reduced cardiac output found, then conduct assesment of volume status.

-  History suggesting fluid loss or hemmorhage
-  Reduced JVP
-  Straight leg raise or fluid challange
-  USG Marker: inferior vena cava collapse, left ventricular stroke volume

reduced cardiac function with increased intravascular volume
-  S3 or S4 gallop
-  JVP increased
-  Extremity Edema
-  Crackles on lung auscultation
-  Chest Xray show cardiomegaly, widening vascular pedicle, kerley B lines, pulmonary edema.
-  ECG: ischemic with or withour chest pain.

If sign of increased cardiac output found, conduct search for cause of reduced SVR.

-  Sepsis
-  Liver Failure
-  Severe Pancreatitis
-  Adrenal Insufficiency
-  Burns
-  Trauma
-  Anaphylaxis
-  Thyrotoxicosis
-  Peripheral AV shunts

The need for arterial line and CVC
- if shock prolonged and doest resolve with proper fluid resuscitation and vasoactive agent.

Initial evaluation followed with early aggresive targeted resuscitation improve survival. If initial bedside evaluation yield confounding data, objective assesment with USG/Echo needed.

### The need for Mechanical Ventilation
Always asses the ability of a patient to protect his or her airway and to maintain adequate gas exhange. Early intubation or mechanical ventilation often required for two main reasons:


1.  Acute hypoxemic respiratory failure.

It may occurs in cardiogenic shock and pulmonary edema, septic shock with pneumonia or acute respiratory distress syndrome (ARDS). 

2.  Ventilatory failure

Often occurs as a consquence of an increased load on the respiratory system in the form of acute metabolic acidosis or decreased lung compliance.

Also in shock, a large percentage of CO need for respiratory muscle (10 folds), lactic acid production from inefficient respiratory muscle activity presents an additional ventilatory load. Ventilatory supports relieve work of breathing and allow redistribution of limited CO to other vital organ.  
Sign of respiratory distress:

-  inability to speak full sentences,
-  accessory muscle activity
-  paradoxical abdominal muscle activity
-  extreme tachypnea (>40 breath/min)

After intubation and mechanical ventilation, declines in MAP frequently seen. The reasons:

1.  Impended Venous Return from positive pressure ventilation (PPV)
2.  reduced endogenous catecholamine secretion once stress associated with respiratory failures abates
3.  Actions of drugs used in endotracheal intubation.
4.  Right heart failure patients or preexisting pulmonary hypertension, due to increased right ventricula afterload due to PPV.

Many patients may be fluid responsive.Therefore, fluid administration and vasopressor support might needed before intubation.

## Respiratory Failure