# Statistics and R languange

## Dataframe Operation
```{r}
worms <- read.csv("mock_data/worms.csv")
```

Checking the columns
```{r}
names(worms)
```

SUmmary
```{r}
summary(worms)
```

The use of aggregates function
```{r}
with(worms, tapply(Worm.density, Vegetation, mean))
```

```{r}
aggregate(worms[,c(2,3,5,7)], list(Community = worms$Vegetation), mean)
```
Multiple classification
```{r}
aggregate(worms[,c(2,3,5,7)],
          list(moisture = worms$Damp, 
               community = worms$Vegetation),
          mean
          )
```

#### Get to Know the Data

##### Check if any unusual data points
```{r}
das <- read.csv("mock_data/das.csv")

plot(das$y)
```

To find which data is the outlier in the above scatter plot
```{r}
which(das$y > 10)
```

So it is the the 50 th data point.


##### Find Relathionsip

###### Numeric vs Numeric Variabales
```{r}
scat <- read.csv("mock_data/scatter.csv")
plot(scat$x, scat$y, pch=21, bg="red")
```

From plot above we can see :
1. The relathioship between response and explanatory variables is curved
2. Degree of scatter from left to right increases (non heterogenous), thus heterocedascity.

###### Categorical vs Numeric Variabales
```{r}
weather <- read.csv("mock_data/weather.data.csv")
plot(factor(weather$month), weather$upper)
```

We see an overall pattern of temperature in a year. We see an outlier at month-6, it turns out a missing data that inputted as zero.

###### making Conditional Plot (COPLOT)

```{r}
coplot <- read.csv("mock_data/coplot.csv")
```

Alter graphics parameter to specifiy two sets of axis on the same row then make a simple scatter plot
```{r}
par(mfrow=c(1,2))
plot(coplot$x, coplot$y)
plot(coplot$y, coplot$x)
```

To look for interaction between explanatory variables and response variables, we employ coplot. It plot Y agains X conditional on the value of Z. Z is numeric variables, by default the coplot split Z into six graphs, with the lowest value appear in the bottom left-hand.
```{r}
coplot(coplot$y ~ coplot$x | coplot$z, pch = 16, panel=panel.smooth)
```
The relathiosip of X and Y change according to Z, as increasing Z the trend comes positive.

###### Interaction in Categorical Data
```{r}
np <- read.csv("mock_data/np.csv")
np$nitrogen <- as.factor(np$nitrogen)
np$phosporus <- as.factor(np$phosphorus)
```

```{r}
par(mfrow=c(1,2))
plot(np$nitrogen, np$yield, main = "Nitrogen")
plot(np$phosporus, np$yield, main = "Phosporus")
```





























